MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-02-17 Article.tex
giga format
better intro + outro

\documentclass[a4paper,num-refs]{oup-contemporary}
% \RequirePackage[2020-02-02]{latexrelease}
% \documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}

\journal{gigascience}

\usepackage{graphicx}
\usepackage{siunitx}

%\usepackage{blindtext}
\usepackage{algorithm}
\usepackage{algcompatible}

\title{MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators}
% \shorttitle{MaBoSS for HPC environments}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1,\authfn{1}]{Adam Šmelko}
\author[2]{Miroslav Kratochvíl}
\author[3,4,5]{Emmanuel Barillot}
% \author[3,4,5]{Laurence Calzone}
\author[3,4,5,\authfn{1}]{Vincent Noël}

\affil[1]{Department of Distributed and Dependable Systems, Charles University, Prague, Czech Republic}
\affil[2]{Luxembourg Centre for Systems Biomedicine, University of Luxembourg, Esch-sur-Alzette, Luxembourg}
\affil[3]{Institut Curie, Université PSL, F-75005, Paris, France}
\affil[4]{INSERM, U900, F-75005, Paris, France}
\affil[5]{Mines ParisTech, Université PSL, F-75005, Paris, France}

\authnote{\authfn{1}smelko@d3s.mff.cuni.cz; vincent.noel@curie.fr}

\papercat{Paper}

\runningauthor{Šmelko et al.}

\jvolume{00}
\jnumber{0}
\jyear{2023}

@@

\begin{frontmatter}
\maketitle

@@

\begin{keywords}
Computational Biology | High Performance Computing | Boolean models
\end{keywords}
\end{frontmatter}

\section{Introduction}
Biological systems are large and complex, and understanding their behavior remains critical to designing new therapies for complex diseases such as cancer. A crucial approach in this endeavor is building computational models from existing knowledge and analyzing them to find intervention points and predict the efficacy of new treatments.
Many different frameworks have been used to describe biological systems, from quantitative systems of differential equations to more qualitative approaches such as Boolean models. While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because it is capable of representing very large systems.
Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to more advanced properties of cancer~\cite{fumia_carcinogenesis_2013, montagud2022prostate}.

@@

\section{Background}

@@

\section{Complexity evaluation and parallelization opportunities of MaBoSS}

@@

\subsection{Simulation complexity and parallelization}

@@

\subsection{Statistics computation complexity and parallelization}

@@

\section{Implementation}\label{sec:implementation}

@@

\subsection{MaBoSS.GPU}
Porting MaBoSS algorithm to GPUs poses an interesting task from the optimizations perspective. A simple rewrite to an annotation-based tool, such as OpenACC, may not be the most optimal strategy due to the usage of complex data structures, such as hash maps and expression trees, that are not well suited for GPU hardware. We divided the implementation into two parts: the bare \emph{simulation} and the trajectory \emph{statistics computation}.

\subsubsection{Simulation}

@@

\subsubsection{Statistics computation}

@@

\subsection{MaBoSS.MPI}
Since MaBoSS.MPI and MaBoSS.GPU projects were developed in parallel, the MPI implementation of MaBoSS is based on the original CPU code. Same as in GPU code, CPU implementation also assigns a single thread to a single trajectory. The MPI implementation simply extends this distribution of work among multiple MPI nodes, adding the second layer of parallelism.

@@

\section{Results}

@@

\subsection{Benchmarking Methodology}

@@

\subsection{MaBoSS.GPU}

@@

\subsection{MaBoSS.MPI}

@@

\section{Conclusions}

@@

\subsection{Future work}

@@

\section{Acknowledgements}
We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.

\section{Funding}
This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773] and SVV project number 260 698.

% \section*{Bibliography}
\bibliography{bibliography}

%% You can use these special %TC: tags to ignore certain parts of the text.
%TC:ignore
%the command above ignores this section for word count
% \onecolumn
% \newpage

>>>>

\RequirePackage[2020-02-02]{latexrelease}
\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
%\usepackage{blindtext}
\usepackage{algorithm}
\usepackage{algcompatible}

% Please give the surname of the lead author for the running footer
\leadauthor{Šmelko}

@@

\title{MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators}
\shorttitle{MaBoSS for HPC environments}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1,\Letter]{Adam Šmelko}
\author[2]{Miroslav Kratochvíl}
\author[3,4,5]{Emmanuel Barillot}
\author[3,4,5]{Laurence Calzone}
\author[3,4,5,\Letter]{Vincent Noël}

\affil[1]{Department of Distributed and Dependable Systems, Charles University, Prague, Czech Republic}
\affil[2]{Luxembourg Centre for Systems Biomedicine, University of Luxembourg, Esch-sur-Alzette, Luxembourg}
\affil[3]{Institut Curie, Université PSL, F-75005, Paris, France}
\affil[4]{INSERM, U900, F-75005, Paris, France}
\affil[5]{Mines ParisTech, Université PSL, F-75005, Paris, France}
\maketitle

@@

\begin{keywords}
Computational Biology | High Performance Computing | Boolean models
\end{keywords}

\begin{corrauthor}
smelko@d3s.mff.cuni.cz, vincent.noel@curie.fr
\end{corrauthor}
\section*{Introduction}
Biological systems are large and complex, and understanding their behavior remains critical to designing new therapies for complex diseases such as cancer. A crucial approach in this endeavor is building computational models from existing knowledge and analyzing them to find intervention points and predict the efficacy of new treatments.
Many different frameworks have been used to describe biological systems, from quantitative systems of differential equations to more qualitative approaches such as Boolean models. While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because it is capable of representing very large systems.
Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to more advanced properties of cancer~\cite{fumia_carcinogenesis_2013, montagud2022prostate}.

@@

\section*{Background}

@@

\section*{Complexity evaluation and parallelization opportunities of MaBoSS}

@@

\subsection*{Simulation complexity and parallelization}

@@

\subsection*{Statistics computation complexity and parallelization}

@@

\section*{Implementation}\label{sec:implementation}

@@

\subsection*{MaBoSS.GPU}
Porting MaBoSS algorithm to GPUs poses an interesting task from the optimizations perspective. A simple rewrite to an annotation-based tool, such as OpenACC, may not be the most optimal strategy due to the usage of complex data structures, such as hash maps and expression trees, that are not well suited for GPU hardware. We divided the implementation into two parts: the bare \emph{simulation} and the trajectory \emph{statistics computation}.

\subsubsection*{Simulation}

@@

\subsubsection*{Statistics computation}

@@

\subsection*{MaBoSS.MPI}
Since MaBoSS.MPI and MaBoSS.GPU projects were developed in parallel, the MPI implementation of MaBoSS is based on the original CPU code. Same as in GPU code, CPU implementation also assigns a single thread to a single trajectory. The MPI implementation simply extends this distribution of work among multiple MPI nodes, adding the second layer of parallelism.

@@

\section*{Results}

@@

\subsubsection*{Benchmarking Methodology}

@@

\subsection*{MaBoSS.GPU}

@@

\subsection*{MaBoSS.MPI}

@@

\section*{Conclusions}

@@

\subsection*{Future work}

@@

\begin{acknowledgements}
We thank Gautier Stoll for his guidance and fruitful discussion.
\end{acknowledgements}

\begin{funding}
This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773] and SVV project number 260 698.
\end{funding}

\section*{Bibliography}
\bibliography{bibliography}

%% You can use these special %TC: tags to ignore certain parts of the text.
%TC:ignore
%the command above ignores this section for word count
\onecolumn
\newpage
